Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages

### 2013-07-28 paper/main.tex
More work on paper.
Rephrased things, updated code snippets to reflect latest code, added motivating example in the intro

\usepackage{amsmath}
\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\WarningsOff

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over the underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without requiring complex proofs to be specified.

@@

An alternative approach is to ensure that a program is semantically sound before it is run. Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled before a program successfully type-checks. A large drawback of this, however, is that automatic verification by the compiler is often impossible, requiring additional proofs to be specified by the developer. This severely limits the uptake of dependently-typed languages.

@@

% Probably could do with rewriting a lot of this
Web applications are also intrinsically side-effecting, requiring extensive network communication, database queries and file I/O. Purely functional programming languages such as Haskell \cite{haskell} handle this monadically, but the composition of different effects can be difficult. Developers either make use of \textit{monad transformers} to compose different effects, or handle everything in a much coarser-grained monad.

\idris{} \cite{brady2011idris} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros. Through the use of \idris{}, and the \texttt{Effects} library \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.

@@

All code used in this paper is available online at \\ \verb+http://www.github.com/SimonJF/IdrisWeb+.
% =================================================
% =================================================

\section{An overview of the \texttt{Effects} library}
\texttt{Effects} \cite{effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity. This section is intended to serve as a brief introduction to the library; for more information, see \cite{brady:effects}.

@@

Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect. Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
{\small
%TODO: Execution
\begin{verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{verbatim}
}
Operations are made available to users by writing functions with the correct \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
{\small
\begin{verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{verbatim}
}
Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context. Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.

@@

\section{Modelling resource usage protocols}
In this section, we show how three such effects; CGI, SQLite and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.

@@

Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved. The only operation that can be performed in this state is initialisation: by calling \texttt{initalise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the \texttt{Cgi} algebraic data type, and involves transitioning from the uninitialised state to the initialised state.

@@

\end{verbatim}
}
User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in turn ensuring that the headers are first written.

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request. Finally, after the headers have been written, the content is then written, which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

@@

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=3em] {\texttt{closeDB}} (A)
        (B) edge[stateEdge, bend left=80]   node[edgeLabel, xshift=3em] {\texttt{connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{stmtFail}} (A)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{finalise}} (B)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
\end{tikzpicture}
}
\caption{SQLite Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

SQLite is a lightweight SQL database engine \cite{sqlite} (TODO: More introduction-y stuff). We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems. Access to the SQLite library is provided through the use of the \idris{} foreign function interface \cite{brady2011idris}, which is beyond the scope of this paper.

@@

\subsection{A Simple Session Handler}
All but the most trivial web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

@@

\section{Type-aware form handling}
Programming web applications often involves processing user data, which may then be used in further effectuful computations. Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.

@@

\subsection{Form Construction}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
{\small
\begin{verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{verbatim}
}
In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We use this to convert between abstract and concrete representations of allowed form types:
{\small
\begin{verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{verbatim}
}
%
In order to specify a form, we once again make use of the \texttt{Effects} library in order to create a resource-aware EDSL. By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form. If a handler function with with different types is supplied, then this will result in a compile-time type error.

@@

{\small
\begin{verbatim}
using (G : Vect FormTy n)
 data FormRes : Vect FormTy n -> Type where
   FR : Nat -> Vect FormTy n -> String -> FormRes G

 data Form : Effect where
   AddTextBox : (label : String) -> 
                (fty : FormTy) -> 
                (Maybe (interpFormTy fty)) -> 
                Form (FormRes G) (FormRes (fty :: G)) () 
   ...
   Submit : interpCheckedFnTy G effs t -> 
            String -> 
            (effs : List WebEffect) -> 
            (t : FormTy) -> 
            Form (FormRes G) (FormRes []) String
\end{verbatim}
}
The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but these are omitted here for brevity.

The resource associated with the EDSL, \texttt{FormRes}, is parameterised over a vector of the types associated with each element. The constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the vector of element types, and the currently generated HTML for the form. It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.

By adding elements to the form, the vector of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}. Additionally, HTML for the form element is generated, and stored in the resource, which is returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.

@@

Before associating a handler function with the form, we must firstly specify the effects available to the handler. This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource. By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form. 
{\small
\begin{verbatim}
useEffects : (effs : List WebEffect) ->
             EffM m [FORM (FormRes G E)] 
                    [FORM (FormRes G effs)] ()
useEffects effs = (UseEffects effs)
\end{verbatim}}
Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}. This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:
\begin{verbatim}
FormHandler : List EFFECT -> Type
FormHandler effs = Eff IO effs ()
\end{verbatim}
In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
%
\begin{verbatim}
  addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
              (fns : HandlerList) ->
              {default tactics { applyTactic 
                findFn 100; solve; }
               prf : FnElem f fns} ->
              EffM m [FORM (FormRes G E)]
                     [FORM (FormRes [] [])] 
                     String
  addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf          
\end{verbatim}
%
Let us look at each aspect of this function in turn. Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function. Note that since we prepend types to the list of \texttt{FormTy}s as opposed to prepending them, we must reverse the list.
{\small
\begin{verbatim}
MkHandlerFnTy : Type
MkHandlerFnTy = (List FormTy, List WebEffect)

@@

       FnHere : {xs : HandlerList, f : 
                 mkHandlerFn ((reverse G), E)} ->
         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
       FnThere : {xs : HandlerList, f : 
                 mkHandlerFn ((reverse G), E)} ->
               FnElem f xs -> FnElem f (x :: xs)
\end{verbatim}
}
\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function. \texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list. %TODO: this should probably be rewritten
We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist. The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function. If the automated proof search fails, compilation will fail with a compile error.

@@

\subsection{Form Handling}
Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation. This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.

@@

\section{Extended Example: Message Board}
Using these components, we may now consider a larger example of a message board, allowing users to register, log in, view and create threads, and list and create new posts in threads.

@@

{\small
\begin{verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{verbatim}
}
The \texttt{handleRequest} function firstly checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function. If so, then the form handling routine is called, which parses the POST data in order to call the correct form handling function. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
{\small
\begin{verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{verbatim}
}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.

{\small
\begin{verbatim}
newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers
\end{verbatim}
}
The form consists of two text boxes: one for the title of the thread, and one for the content of the first post. Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value. The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of form handlers, and takes the following form:
{\small
\begin{verbatim}
handlers : HandlerList
handlers = [(([FormString, FormString], 
              [CgiEffect, SessionEffect, SqliteEffect]) ** 
               (handleRegisterForm, "handleRegisterForm")),

@@

We believe we have demonstrated that embedded domain-specific languages can be used to make use of the power of dependently-typed languages without incurring the traditional overheads of 
\subsection{Further Work}
We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.

@@

% We recommend abbrvnat bibliography style.
% #firstworldanarchists
\bibliographystyle{plainnat}

>>>>

\usepackage{amsmath}
\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usetikzlibrary{arrows,automata}
\WarningsOff

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.

@@

%%------- EXAMPLE
Allow us to consider a simple, traditional database access routine, written in PHP, where we wish to obtain the name and address of every employee working in a given department, \texttt{\$dept}. We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:

{\small
\begin{verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{verbatim}
}

We must subsequently check to see if the connection was successful, and cease execution of the database routine if so. Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.

{\small
\begin{verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{verbatim}
}

We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:

{\small
\begin{verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param($dept);
\end{verbatim}
}

After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn. At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:

{\small
\begin{verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);

  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{verbatim}
}

Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:

{\small
\begin{verbatim}
  $stmt->close();
  $conn->close();
\end{verbatim}
}

Even in this small example, there exists a very specific resource usage protocol. Firstly, a connection to the database must be opened. The object-oriented style used in the example encapsulates this to an extent, as the object must be created in order for operations to be performed, however it is less obvious in a procedural version of the code. Secondly, a prepared statement is created, using the raw SQL and placeholders to which variables are later bound. The statement is then executed, and each row is retrieved from the database. Finally, the resources are freed.

There is, however, no way to statically guarantee that this protocol is adhered to. A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error. Similarly, a programmer may omit closing the statement or connection, and can lead to substantial problems such as handle leaks in longer-running server applications.

An alternative approach is to ensure that a program is semantically sound before it is run. Through the use of \textit{dependent types}, it is possible to impose arbitrarily strong specifications on programs, requiring that these specifications are fulfilled prior to being accepted by a type checker. A large drawback of this, however, is that automatic verification by the compiler is often impossible, requiring additional proofs to be specified by the developer. This severely limits the uptake of dependently-typed languages in real-world settings.

@@

\idris{} \cite{brady2011idris} is a language with full dependent types, and extensive support for embedded domain-specific languages through ad-hoc syntax overloading and syntax macros. Through the use of \idris{}, and the \texttt{Effects} framework \cite{brady:effects}, we present a dependently-typed web framework, which allows for the construction of programs with additional guarantees about correctness and security, whilst minimising the increase in development complexity.

@@

We structure the remainder of this paper as follows. We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocol for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.

The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
% =================================================
% =================================================

\section{An overview of the \texttt{Effects} framework}
\label{effects}
\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity. This section is intended to serve as a brief introduction to the framework; for more information, see \cite{brady:effects}.

@@

Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect. Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
{\small
%TODO: Execution
\begin{verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{verbatim}
}
Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
{\small
\begin{verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{verbatim}
}
Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context. Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.

@@

\section{Modelling resource usage protocols}
\label{rup}
In this section, we show how three such effects; CGI, SQLite and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.

@@

Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved. The only operation that can be performed in this state is initialisation: by calling \texttt{initalise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.

@@

\end{verbatim}
}
User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in accordance with the resource usage protocol.

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request. Finally, the headers and content are written which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

@@

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
\end{tikzpicture}
}
\caption{SQLite Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications. We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems. Access to the SQLite library is provided through the use of the \idris{} foreign function interface \cite{brady2011idris}, which is beyond the scope of this paper.

@@

\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

@@

\section{Type-aware form handling}
\label{form}
Programming web applications often involves processing user data, which may then be used in further effectuful computations. Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.

@@

\subsection{Form Construction}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
{\small
\begin{verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{verbatim}
}
In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We use this to convert between abstract and concrete representations of allowed form types:
{\small
\begin{verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{verbatim}
}
%
In order to specify a form, we once again make use of the \texttt{Effects} library in order to create a resource-aware EDSL. By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.

@@

{\small
\begin{verbatim}
using (G : List FormTy, E : List WebEffect)
  data FormRes : List FormTy -> 
                 List WebEffect -> Type where
    FR : Nat -> 
         List FormTy -> 
         List WebEffect -> 
         String -> 
         FormRes G E

  data Form : Effect where
    AddTextBox : (label : String) -> 
                 (fty : FormTy) -> 
                 (Maybe (interpFormTy fty)) -> 
                 Form (FormRes G E) 
                      (FormRes (fty :: G) E) () 
   ...

    Submit : (mkHandlerFn ((reverse G), E)) ->
             String -> 
             Form (FormRes G E) (FormRes [] []) String
\end{verbatim}
}
The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but these are omitted here for brevity.

The resource associated with the EDSL, \texttt{FormRes}, is parameterised over a list of the types associated with each element and the effects supported by the handler function. The constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form. It is necessary to keep track of the element types at both the type and value level as we must use the values in later computations when serialising the handler function.

By adding elements to the form, the list of form types \texttt{G} is updated, as seen in the output value of \texttt{AddTextBox}. Additionally, HTML for the form element is generated, and stored in the resource, which is returned by the \texttt{addSubmit} function in order for it to be displayed on the web page.

@@

Before associating a handler function with the form, we must firstly specify the effects available to the handler. This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource. By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form. 
{\small
\begin{verbatim}
  useEffects : (effs : List WebEffect) ->
               EffM m [FORM (FormRes G E)] 
                      [FORM (FormRes G effs)] ()
  useEffects effs = (UseEffects effs)
\end{verbatim}}
Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}. This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:
{\small
\begin{verbatim}
  FormHandler : List EFFECT -> Type
  FormHandler effs = Eff IO effs ()
\end{verbatim}
}
In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
{\small
\begin{verbatim}
  addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
              (fns : HandlerList) ->
              {default tactics { applyTactic 
                findFn 100; solve; }
               prf : FnElem f fns} ->
              EffM m [FORM (FormRes G E)]
                     [FORM (FormRes [] [])] 
                     String
  addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf          
\end{verbatim}
}
Let us look at each aspect of this function in turn. Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function. Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
{\small
\begin{verbatim}
MkHandlerFnTy : Type
MkHandlerFnTy = (List FormTy, List WebEffect)

@@

       FnHere : {xs : HandlerList, f : 
                 mkHandlerFn ((reverse G), E)} ->
         FnElem f ((((reverse G), E) ** (f, fStr)) :: xs)
       FnThere : {xs : HandlerList, f : 
                 mkHandlerFn ((reverse G), E)} ->
               FnElem f xs -> FnElem f (x :: xs)
\end{verbatim}
}
\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the form elements and the effects used by the handler function. \texttt{FnHere} demonstrates that the element is at the head of the current point of the list, whereas \texttt{FnThere} demonstrates that the element is at some point further in the list. %TODO: this should probably be rewritten
We may then use linguistic reflection and a simple automated proof search to automatically generate the proof at compile time, should one exist. The proof may then be used in subsequent computations: in our case, we use it to retrieve the unique identifier for the function. If the automated proof search fails, compilation will fail.

@@

\subsection{Form Handling}
\label{formhandling}
Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation. This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.

@@

\section{Extended Example: Message Board}
\label{messageboard}
Using these components, we may now consider a larger example of a message board, allowing users to register, log in, view and create threads, and list and create new posts in threads.

@@

{\small
\begin{verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{verbatim}
}
The \texttt{handleRequest} function firstly checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function. If so, then the form handling routine is called, which handles the form data as described in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
{\small
\begin{verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{verbatim}
}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.

{\small
\begin{verbatim}
newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers
\end{verbatim}
}
The form consists of two text boxes: one for the title of the thread, and one for the content of the first post. Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value. The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
{\small
\begin{verbatim}
handlers : HandlerList
handlers = [(([FormString, FormString], 
              [CgiEffect, SessionEffect, SqliteEffect]) ** 
               (handleRegisterForm, "handleRegisterForm")),

@@

\subsection{Further Work}
We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.

@@

% We recommend abbrvnat bibliography style.

\bibliographystyle{plainnat}
